---
title: "Biasfile_1"
output: html_document
---

This R script creates a biasfile for use with Maxent by creating a density layer using the kde2d function, it also includes code to run Maxent within R, but the current version of dismo does not allow for Maxent to be run within R while factoring a biasfile.   

```{r}
Sys.setenv(JAVA_HOME = '/Library/Java/JavaVirtualMachines/jdk-11.0.15.1.jdk/Contents/Home')
library(dismo) # interface with MaxEnt
library(raster) # spatial data manipulation
library(MASS) # for 2D kernel density function
library(magrittr) # for piping functionality, i.e., %>%
library(maptools) # reading shapefiles
library(rgeos)
library(rJava)
library(data.table) #read in location data easier from raw ALA downloads
```
  
Read in occurence data
```{r}
locations <- fread("passerine_records_1970_onwards.csv", select = c("decimalLongitude", "decimalLatitude"))
```

Clean up reformat data
```{r}
locations <- na.omit(locations)
colnames(locations) <- c("lon", "lat") #rename colnames for rasterize function 
climdat <- brick("~/MaxEnt/1960-1979EnvData/ANUCLIM19601979maxt_mean.asc") #use arbitrary climate data file as a template for rasterize
occur.ras <- rasterize(locations, climdat, 1) 
plot(occur.ras)
```

Crop to Australia and reformat for kde2d function 
```{r}
extent_aus <- c(110, 158, -46, -9)
occur.ras<- crop(occur.ras, extent_aus)

presences <- which(values(occur.ras) == 1) 
pres.locs <- coordinates(occur.ras)[presences, ]
```

Create biasfile
```{r}

dens <-kde2d(pres.locs[,1], pres.locs[,2],
             n = c(nrow(occur.ras), ncol(occur.ras)))
dens %>%
  raster %>%
  resample(climdat, method="bilinear") %>% 
  crop(extent_aus) %>%
  writeRaster('biasfile.asc')

```

Run Maxent (currently does not work)
```{r}
biasfile <- raster("biasfile.asc")
mod1 <- maxent(b1, locations, args = "biasfile=biasfile")
```

