---
title: "PrepDataRunMaxent"
output: html_document
date: "2022-09-02"
---
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '~/Honours')
```


CLIMATE
Load in packages and set up variables 
```{r}
library(ncdf4) #package for opening .nc files
library(terra) # package for spatraster manipulation
library(rgdal) # package for geospatial analysis
library(magrittr) #package for piping
library(data.table) #package for easy reading in csvs
library(climates) #package for making bioclim variables
library(ntbox) #package for running maxent
library(stringr) #package for manipulating strings

varlist <- c('tavg', 'tmax', 'tmin', 'rain')
extent_aus <- ext(112, 154, -44,-9)
months <- c("01", "02", "03", "04", "05", "06", "07", "08", "09", 10:12) #vector of months
yearlist <- c(1980:2015) #vector of years


```

Functions to download and open files from database for ANUCLIM data
```{r}

nc.open <- function(month, variable, year) { #Function to get raster with inputs that can be iterated
  a = nc_open(paste0("https://dapds00.nci.org.au/thredds/dodsC/gh70/ANUClimate/v2-0/stable/month/", variable,"/", year, "/ANUClimate_v2-0_", variable, "_monthly_", year, month,".nc"))
  b = ncvar_get(a, variable) %>% t() %>% rast(crs = "WGS84") %>% set.ext(extent_aus) %>% as.data.frame()
  return(b)
}

nc.open.xy <- function(month, variable, year) { #Function to get raster with inputs that can be iterated
  a = nc_open(paste0("https://dapds00.nci.org.au/thredds/dodsC/gh70/ANUClimate/v2-0/stable/month/", variable,"/", year, "/ANUClimate_v2-0_", variable, "_monthly_", year, month,".nc"))
  b = ncvar_get(a, variable) %>% t() %>% rast(crs = "WGS84") %>% set.ext(extent_aus) %>% as.data.frame(xy = T)
  return(b)
}
```

Make object for list of coordinates of all data
``` {r}
bioclim.xy <- nc.open.xy("01", "tmax", 1960) 
bioclim.xy <- cbind(bioclim.xy$x, bioclim.xy$y) 
colnames(bioclim.xy) <- c("lon", "lat")
```

Loop to download and process ANUCLIM data into bioclim variables as raster layers 
```{r}
for (i in yearlist) {
  bioclimdf.list = list()
  for (ii in varlist) {
    temp.df = data.frame(matrix(nrow=7050282, ncol=0))
    temp.df = lapply(months, nc.open, variable = ii, year = i) %>%
      bind_cols()
    bioclimdf.list[[ii]] = temp.df
  }
  bioclim <- bioclim2(bioclimdf.list$tmin, bioclimdf.list$tmax, bioclimdf.list$rain, bioclimdf.list$tavg, files.as.inputs = F)
  bioclim.dataframe <- cbind(bioclim.xy, bioclim)
  bioclim.rast <- rast(bioclim.dataframe, type = "xyz")
  bioclim.rast[is.na(bioclim.rast)] <- -9999
  names.list <- names(bioclim.rast)
}

# Write to disk
for (iii in names.list) {
  writeRaster(subset(bioclim1.rast, iii), paste0("data/proj_layers/temp/", iii,".asc"), NAflag = -9999, overwrite = T)
}
```

Calculate means for years 1960-1979 to make the base maxent model (climate)
Set up environment
```{r}
bioclim.list <- c('tavg', 'tmax', 'tmin', 'rain')
years <- c(1960:1979) 

nc.open <- function(year, month, variable) { #Function to get raster with inputs that can be iterated
  a = nc_open(paste0("https://dapds00.nci.org.au/thredds/dodsC/gh70/ANUClimate/v2-0/stable/month/", variable,"/", year, "/ANUClimate_v2-0_", variable, "_monthly_", year, month,".nc"))
  b = ncvar_get(a, variable) %>% t() %>% rast(crs = "WGS84") %>% set.ext(extent_aus)
  b[is.na(b)] <- -9999
  NAflag(b) <- -9999
  return(b)
}
```

Loop to create dataframes for input variables, function "bioclim2" takes dataframes as inputs
```{r}
for (i in bioclim.list) { #for each input variable (tavg, tmas, tmin, rain)
  clim.dataframe <- data.frame(matrix(nrow=7050282, ncol=0)) #make empty dataframe 
  for (ii in months){
    clim.dataframe = lapply(years, nc.open, month = ii, variable = i) %>%
    rast() %>% #make rasters of all 
    mean() %>% #average over all months of the year
    as.data.frame() %>% #coerce to dataframe
    cbind(clim.dataframe) #add to dataframe
  }
  colnames(clim.dataframe) <- months
  write.csv(clim.dataframe, paste0("BioClim_dataframes/", i,".csv")) #write to disk
}

```

Read in input dataframes
```{r}
bioclim <- bioclim2(tmin = fread("data/clim_data/BioClim_dataframes/tmin.csv", select = c(2:13)), tmax = fread("data/clim_data/BioClim_dataframes/tmax.csv", select = c(2:13)), prec = fread("data/clim_data/BioClim_dataframes/rain.csv", select = c(2:13)), tmean = fread("data/clim_data/BioClim_dataframes/tavg.csv", select = c(2:13)), cov = F, files.as.inputs = F)

```

Convert dataframe to raster and write into dirctory 
```{r}
bioclim.dataframe <- cbind(bioclim.xy, bioclim) #give coordinates
bioclim.rast <- rast(bioclim.dataframe, type = "xyz") #convert results of bioclim2 function to raster
bioclim.rast[is.na(bioclim.rast)] <- -9999
names.list <- names(bioclim.rast)

for (i in names.list) {
  writeRaster(subset(bioclim.rast, i), paste0("~/Honours/data/clim_data/ANUBIOCLIM_ascs/", i,".asc"), NAflag = -9999, overwrite = T)
}
```

LANDCOVER

Set up environment
```{r}
ld_states <- nc_open("data/land_data/LUH2_GCB2019_states.nc4") #load in .nc file with landcover data

#Set extents
lon <- ncvar_get(ld_statesA, "lon")
lon[lon > 180] <- lon[lon > 180] - 360
lat <- ncvar_get(ld_statesA, "lat", verbose = F)
t <- ncvar_get(ld_statesA, "time")
extent_aus <- ext(110, 158, -46, -9) #different from above because of  cropping needs
world = ext(min(lon), max(lon), min(lat), max(lat))
temp <- rast("/Users/u6638201/MaxEnt/biasfile.asc") #use as template for resample
varlist <- c("pastr", "primf", "primn", "range", "secdf", "secdn", "secmb", "urban")

```

Function to retrieve rasterlayers from array and write to disk
```{r}
get.raster <- function(x, y, z) {
    a = x - 849
    b = y[a,,] %>% t() %>% rast(crs = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs+ towgs84=0,0,0") %>% flip(direction='v') %>% set.ext(world) %>% crop(extent_aus)  %>% resample(temp, method = "rms") 
    b[is.na(b)] <- -9999
    writeRaster(b, paste0("data/land_data/1980-2015_ascs/", x,"/", z,".asc"), overwrite = T, NAflag = -9999)
}
```

Loop to make layers
```{r}
for (i in varlist) {
  array = ncvar_get(ld_states, i)
  lapply(yearlist, get.raster, y = array, z = i)
}
```

Calculate means for years 1960-1979 to make the base maxent model (landcover)

Function to calculate means
```{r}
land.mean <- function(x, y, z) {
  a = x - 849
  b = y[a,,] %>% t() %>% rast(crs = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs+ towgs84=0,0,0") %>% flip(direction='v') %>% set.ext(world) %>% resample(temp, method = "rms") 
  b[is.na(b)] <- -9999
  NAflag(b) <- -9999
  return(b)
}
```

Loop to make layers and write to disk
```{r}
for (i in varlist) {
  array = ncvar_get(ld_states, i)
  rast.list = lapply(years, land.mean, y = array, z = i) #using years object from earlier
  b = rast(rast.list) %>% mean() 
  b[is.na(b)] <- -9999
  writeRaster(b, paste0("data/land_data/land_layers/", i, ".asc"), overwrite = T, NAflag = -9999)
}
```

MAXENT
(see vscode for solution that was used for climate layers)
Run maxent for landcover
```{r}
for (i in yearlist){
  maxent_call(
    "~/Desktop/maxent-2",
    features = c("h","p"),
    environmentallayers = "data/land_data/land_layers",
    samplesfile = "data/bird_data/combined_1960_1979_30km_thin1.csv",
    outputdirectory = "outputs/proj_layers/temp/",
    projectionlayers = paste0("data/land_data/1980-2015_ascs/",i),
    extrapolate = T,
    askoverwrite = T,
    randomtestpoints = 25,
    betamultiplier = 0.5,
    biasfile = "data/biasfile.asc",
    doclamp = T,
    randomseed = T,
    threads = 6)
  
  rast(paste0("outputs/proj_layers/temp/land_layers_bmult_0.5_hp/Malurus_", i,".asc")) %>%
    writeRaster(paste0("outputs/land_max_proj_norep/", i, ".tif"), overwrite = T)
  unlink("outputs/proj_layers/temp/land_layers_bmult_0.5_hp", recursive = T)
}
```

Extract Data for Bird Samples

Climate

Read in projections in raster format for 1980 - 2015
```{r}
output.file.list <- list.files("outputs/proj_layers/years", full.names = T) 
output.rasters <- list() #make empty list  

for (i in output.file.list) { #fill with rasters of maxent projections
  output.rasters = c(output.rasters, rast(i))
}

names(output.rasters) <- c(1980:2020) #set raster names so they can be indexed

```

Read in sample ID's and locations
```{r}
bird.data <- read.csv("data/bird_data/Honours Thesis Metadata v2.csv") #Read in sample metadata 
bird.data$Year <- str_sub(bird.data$Date, start = -4) #get years from date format and write to column
bird.data$Year[bird.data$Year == ""] = NA
bird.data.trimmed <- cbind(bird.data$EXP.ID, bird.data$Longitude, bird.data$Latitude, bird.data$Year) %>% 
  na.omit() %>% #Trim data down to just ID's and Locations 
  as.data.frame()
```

Extract points from projected layers by sample ID and collection year
```{r}
bird.id <- bird.data.trimmed$V1 #Make list of sample ID's to index data by
bird_values <- c() 

for (i in bird.id) {
  year = bird.data.trimmed$V4[bird.data.trimmed$V1==i]
  xy = matrix(c(as.numeric(bird.data.trimmed$V2[bird.data.trimmed$V1==i]), 
                as.numeric(bird.data.trimmed$V3[bird.data.trimmed$V1==i])), ncol = 2)
  bird_values = extract(output.rasters[[year]], xy) %>% 
    c(bird_values, .) 
}

bird_values <- bird_values %>% unlist() %>% data.frame() #Transform to dataframe
rownames(bird_values) <- bird.id

bird_values <- cbind(bird.data.trimmed, bird_values) %>% na.omit #Add metadata
write.csv(bird_values, "outputs/sampleClimMaxentValues1980_2020.csv") #Write to disk
```

Extract for landcover 

Read in projections in raster format for 1980 - 2015
```{r}
output.file.list <- list.files("outputs/land_max_proj_norep", full.names = T) 
output.rasters <- c() # make empty list  

for (i in output.file.list) { #fill with rasters of maxent projections
  output.rasters = c(output.rasters, rast(i))
}

```

Extract points from projected layers by sample ID and collection year
```{r}
bird.id <- bird.data.trimmed$V1 #Make list of sample ID's to index data by
bird_values <- c() 

for (i in bird.id) {
  year = bird.data.trimmed$V4[bird.data.trimmed$V1==i]
  xy = matrix(c(as.numeric(bird.data.trimmed$V2[bird.data.trimmed$V1==i]), 
                as.numeric(bird.data.trimmed$V3[bird.data.trimmed$V1==i])), ncol = 2)
  bird_values = extract(output.rasters[[year]], xy) %>% 
    c(bird_values, .) 
}

bird_values <- bird_values %>% unlist() %>% data.frame() #Transform to dataframe
rownames(bird_values) <- bird.id

bird_values <- cbind(bird.data.trimmed, bird_values) %>% na.omit #Add metadata
write.csv(bird_values, "outputs/sampleLandMaxentValues1980_2020.csv") #Write to disk
```

